Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-14 Dail_Mad_ext.tex
edits to results and discussion

We used the \texttt{unmarked} library in \textbf{R} to estimate the parameters 
for each simulation with the same
initial abundance (Poisson) and dynamics models as were simulated.
When run in a maximum likelihood framework, these models require a
a finite upper bound $N_{max}$,
\citep{royle:2004biom,dail_madsen:2011};
%(Royle 2004, Dail and Madsen 2011);
we used 200.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95\% confidence
intervals for parameters that overlap the true values).

@@

We applied our model extensions to North American Breeding Bird Survey
(BBS) data collected from 1966-2010 in the bordering US states
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
forest-breeding migratory songbird with a stable or increasing trend
in the region \citep{porneluzi_etal:2011}. 
%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
%only breeds in the western parts of these states and has been declining in the
%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

%GENERAL ABOUT BBS DATA.
The BBS is an annual roadside survey implemented by trained
observers in the United States and Canada. An observer conducts 50
3-minute point counts, with 400 m radii and 0.8 km apart from each other along a
39.4 km route. We only used data marked as acceptable for use in the annual BBS
analysis \citep{sauer_etal:1994auk}.  We summed the number of ovenbirds
seen on each route and year, and used the routes (rather
than the individual stops) as our sites.
Strong winds can interfere with point count observers' ability
to hear birds \citep{simons_etal:2007}; we tested the effects of wind
speed on detection probability, using the truncated mean of wind conditions at the start and end of
each route on the Beaufort scale as a categorical predictor. 
%BBS volunteers record wind conditions
%at the beginning and end of each route on the Beaufort Scale
%\citep[start and end wind]{robbins_etal:1986}.
%%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
%When start or end wind was not recorded we imputed those values with
%the start or end mean. We took the mean of start and end wind for
%each route and put this average wind scale value into four
%categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
%$\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an observer ran a route as a predictor variable for detection
probability.

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with non-informative priors.  
% JEFF: We need to be specific about the priors
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.

@@

Estimator bias and error were generally low with the exponential model, and 
coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table S2.1 in Appendix 2). 
However, estimates of $r$ were 
imprecise (RMSE between 0.020 and 0.025; relative RMSE between 413\% and 
482\% for non-zero true $r$) and biased low (between -0.008 and -0.005; relative bias between
-151\% and -101\% for non-zero true $r$) when the true value of $\Lambda$ was low (1). 
Simulations of the Ricker-logistic model also generally performed well (Fig.~\ref{fig:rick_hists}; 
Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
biased high (between 65.4\% and 198.9\%), and had low coverage (between 77.6\% and 94.3\%).
In other cases for the Ricker-logistic model, relative RMSE was between 6.7\%
and 55\%, bias was between -6.4\% and 5.8\%, and coverage was between
91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
performed fairly well (Fig.~\ref{fig:ricki_hists}; Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
93.2\% and 96.4\%).

@@

Estimates for the top ranked model %for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
%(to 42.5 $\pm$ 7.3 and 101.8 $\pm$ 21.1,
%respectively),
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1 (on the logit scale), and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

We estimated ovenbird abundance by route and year (and averaged over routes) 
using the Bayesian model with random observer effects (Fig.~\ref{fig:oven_N}).  
Estimated average abundance over time alternated between growing and stable periods.  
Estimated ovenbird
abundance was highest in western and eastern Maryland in 1970 and 1980, but became
more evenly distributed by 2010.  
%These aren't up to date, and it doesn't look like people use CV as a measure of 
%uncertainty in MCMC estimates anyway.
%Coefficient of variation (CV) for route and year specific
%abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
%for the estimate of mean initial abundance ($\Lambda$) was 0.15.
As expected, uncertainty in estimates of route and year specific abundance was high.

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

>>>>

We used the \textbf{R} package \texttt{unmarked} to estimate the parameters 
for each simulation with the same
initial abundance (Poisson) and dynamics models as were simulated.
When run in a maximum likelihood framework, these models require a
a finite upper bound $N_{max}$,
\citep{royle:2004biom,dail_madsen:2011};
%(Royle 2004, Dail and Madsen 2011);
we used 200.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95\% confidence
intervals for parameters that overlap the true values).

@@

We applied our model extensions to North American Breeding Bird Survey
(BBS) data collected from 1966-2010 in %the bordering US states
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
forest-breeding migratory songbird with a stable or increasing trend
in the region \citep{porneluzi_etal:2011}. 
%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
%only breeds in the western parts of these states and has been declining in the
%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

%GENERAL ABOUT BBS DATA.
The BBS is an annual roadside survey implemented by trained
observers in the United States and Canada. An observer conducts 50
3-minute point counts, with 400 m radii and 0.8 km apart 
%from each other 
along a
39.4 km route. We only used data marked as acceptable for use in the annual BBS
analysis \citep{sauer_etal:1994auk}.  We summed the number of ovenbirds
seen on each route and year, and used the routes (rather
than the individual stops) as our sites.
Strong winds can interfere with point count observers' ability
to hear birds \citep{simons_etal:2007}; we tested the effects of wind
speed on detection probability, using the truncated mean of wind conditions at the start and end of
each route on the Beaufort scale as a categorical predictor. 
%BBS volunteers record wind conditions
%at the beginning and end of each route on the Beaufort Scale
%\citep[start and end wind]{robbins_etal:1986}.
%%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
%When start or end wind was not recorded we imputed those values with
%the start or end mean. We took the mean of start and end wind for
%each route and put this average wind scale value into four
%categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
%$\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an observer ran a route as a predictor variable for detection
probability.

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with non-informative priors.  
% JEFF: We need to be specific about the priors
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we used %obtained these estimates from 
a single chain.  %We present these estimates for ovenbirds only.

@@

Estimator bias and error were generally low with the exponential model, and 
coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table S2.1 in Appendix 2). 
However, estimates of $r$ were 
imprecise (RMSE between 0.020 and 0.025; relative RMSE between 413\% and 
482\% for non-zero true $r$) and biased low (between -0.008 and -0.005; relative bias between
-151\% and -101\% for non-zero true $r$) when the true value of
$\Lambda$ was low. % (1). 
Simulations of the Ricker-logistic model %also 
generally performed well (Fig.~\ref{fig:rick_hists}; 
Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
biased high (between 65.4\% and 198.9\%), and had low coverage (between 77.6\% and 94.3\%).
In other cases for the Ricker-logistic model, relative RMSE was between 6.7\%
and 55\%, bias was between -6.4\% and 5.8\%, and coverage was between
91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
performed fairly well (Fig.~\ref{fig:ricki_hists}; Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
93.2\% and 96.4\%).

@@

Estimates for the top ranked model %for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved 
%% JEFF: How do we explain that????
(Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
%(to 42.5 $\pm$ 7.3 and 101.8 $\pm$ 21.1,
%respectively),
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1 (on the logit scale), and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

We estimated ovenbird abundance by route and year (and averaged over routes) 
using the Bayesian model with random observer effects (Fig.~\ref{fig:oven_N}).  
Estimated average abundance over time alternated between growing and stable periods.  
Estimated ovenbird
abundance was highest in western and eastern Maryland in 1970 and 1980, but became
more evenly distributed by 2010.  
%These aren't up to date, and it doesn't look like people use CV as a measure of 
%uncertainty in MCMC estimates anyway.
%Coefficient of variation (CV) for route and year specific
%abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
%for the estimate of mean initial abundance ($\Lambda$) was 0.15.
%As expected, uncertainty in estimates of route and year specific abundance was high.

@@

Our work %has highlighted the limitations of classical state-space
%models as applied to ecological time series data, and we have
%demonstrated 
demonstrates how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes 
%these limitations. 
many of the limitations of classical state-space models as applied to
ecological time series data.
We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.
